A flexible Prolog interpreter in Python

### 2012-08-17 talk/badhonnef2007/talk.tex
- repetition - typo
Otherwise the talk looks very good to me.
tune plot

  \setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}

@@

\begin{frame}
  \frametitle{Outline}
  \tableofcontents
  % You might wish to add the option [pausesections]
\end{frame}

@@

  \begin{itemize}
  \item
    Pyrolog is a Prolog interpreter written in RPython
  \item
    RPython (``Restricted Python'') is a subset of Python translatable to other
    languages
  \item
    RPython is designed to be significantly faster than regular Python
  \item
    translation tool-chain part of the PyPy project
  \end{itemize}
\end{frame}

\section{The PyPy Approach to VM Construction}
\subsection{Overview}
\begin{frame}
  \frametitle{What is PyPy?}
  \begin{itemize}
  \item
    started as a Python VM implementation in Python
  \item
    Python itself is too dynamic to be translatable to other languages, need a subset
  \item
    includes a translation tool-chain for RPython
  \item
    is becoming a general environment for writing interpreters (JavaScript, Prolog started)
  \item
    Open Source project (MIT license)
  \item
    received EU funding for 2.5 years
  \end{itemize}

@@

\begin{frame}
  \frametitle{The Python case (i)}
  CPython (the reference implementation) is a straightforward, portable VM.

  \begin{itemize}
  \item
    Pervasive decisions: reference counting, global lock \dots
  \item
    No dynamic compilation
%  \pause
  \end{itemize}
  \begin{block}{
    Extensions:}
    \begin{itemize}
    \item
      \alert{Stackless} (unlimited recursion, coroutines, serializable continuations)
    \item
      \alert{Psyco} (run-time specializer)
    \item
      \alert{Jython}, \alert{IronPython}
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The Python case (ii)}
  \begin{block}{Problems of Extensions:}
    \begin{itemize}
    \item hard to maintain: need to keep track of CPython
    \item Psyco very hard to port to other hardware architectures
    \item tedious to write them, Python semantics need to be re-implemented
    \end{itemize}
  \end{block}
  \begin{itemize}
  \item
    The community wants Python to run everywhere:
    Jython (Java), IronPython (.NET).
    Lots of effort and duplication.

  \item
    At various points various incompatibilities between
    the implementations

  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{The Prolog case (ii)}
  \begin{block}{mature C implementations}
    \begin{itemize}
    \item
      interfacing with libraries is tedious
    \item
      changing the language to experiment is hard
    \item
      often extensions to core Prolog, incompatible between each other
    \item
      fixed implementation decisions (GC, how to generate code, etc.)
    \end{itemize}
  \end{block}
  \begin{block}{implementations on CLR and JVM}
    \begin{itemize}
    \item
      interfacing with libraries of the platform mostly easy
    \item
      no extensions to core Prolog (like tabling, coroutines)
    \item
      slow, compared to good C implementations
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Mapping to low-level targets}
  \begin{itemize}
  \item
    Mechanically translate the interpreter to multiple
    lower-level targets
    \begin{itemize}
    \item C-like
    \item Java
    \item .NET
    \end{itemize}
  \item
    Insert low-level aspects into the code as required by
    the target (Object layout, memory management)
    \begin{itemize}
    \item object layout
    \item memory management
    \end{itemize}  \item
    Optionally insert new pervasive features not expressed
    in the source
    \begin{itemize}
    \item continuations, ``micro-threads''
    \item dynamic compilation
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Translation Aspects (i)}
  Features not present in the source can be added during translation.
  \begin{itemize}
  \item
    \alert{memory management}: use different GC strategies (Boehm collector, custom mark-n-sweep)
  \item
    \alert{Stackless transformation}: allows program to control its stack (continuations, \dots)
  \end{itemize}
\end{frame}

@@

\section{The Prolog interpreter}
\subsection{Interpreter}
\begin{frame}
  \frametitle{Prolog Interpreter Implementation}
  \begin{itemize}
  \item
    naive, very simple interpreter
    \begin{itemize}
    \item uses "structure copying"
    \item interprets Prolog terms directly, no bytecode
    \end{itemize}
  \item
    uses continuation passing style inspired by BinProlog
  \item
    Prolog calls mapped to RPython calls
  \item
    implements large parts of the ISO standard (some builtins missing)
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Interpreter Facts}
  \begin{itemize}
  \item
    2500 lines of Python code in total
  \item
    700 of those are for builtins
  \item
    after translation to C: 14000 line of C code
  \item
    part of the PyPy distribution at:
    \texttt{http://codespeak.net/pypy}
  \end{itemize}
\end{frame}

@@

  % Keep the summary *very short*.
  \begin{itemize}
  \item
    Very simple Prolog interpreter in RPython can compete with interpreters on the JVM, CLR
  \item
    Interpreter implementation eased by use of a high-level language
  \item
    Low-level details abstracted away but re-inserted later
  \end{itemize}
\end{frame}

@@

>>>>

  %\setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}

@@

%\begin{frame}
%  \frametitle{Outline}
%  \tableofcontents
  % You might wish to add the option [pausesections]
%\end{frame}

@@

  \begin{itemize}
  \item
    Pyrolog is a Prolog interpreter written in Python
  \item
    want to compile it to get interesting performance
  \item
    translation tool-chain part of the PyPy project
  \item
    Python itself is too dynamic to be translatable to other languages, need a subset
  \item
    RPython (``Restricted Python'') is a subset of Python translatable to other
    languages
  \item
    RPython is designed to be significantly faster than regular Python
  \end{itemize}
\end{frame}

\section{The PyPy Approach to VM Construction}
\subsection{Overview}
\begin{frame}
  \frametitle{What is PyPy?}
  \begin{itemize}
  \item
    started as a Python VM implementation in Python
  \item
    most important part: translation tool-chain for RPython
  \item
    is becoming a general environment for writing interpreters (JavaScript, Prolog started)
  \item
    Open Source project (MIT license)
  \item
    received EU funding for 2.5 years
  \end{itemize}

@@

\begin{frame}
  \frametitle{The Python case}
  CPython (the reference implementation) is a straightforward, portable VM.

  \begin{itemize}
  \item
    Pervasive decisions: reference counting, global lock \dots
  \item
    No dynamic compilation
  \end{itemize}
  \pause
  \begin{block}{
    Extensions:}
    \begin{itemize}
    \item
      \alert{Stackless} (unlimited recursion, coroutines, green threads)
    \item
      \alert{Psyco} (run-time specializing compiler)
    \item
      \alert{Jython}, \alert{IronPython}
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{The Prolog case (ii)}
  \begin{block}{mature C implementations}
    \begin{itemize}
    \item
      interfacing with libraries is tedious
    \item
      changing the language to experiment is hard
    \item
      often extensions to core Prolog, incompatible between each other
    \end{itemize}
  \end{block}
  \pause
  \begin{block}{implementations on CLR and JVM}
    \begin{itemize}
    \item
      interfacing with libraries of the platform mostly easy
    \item
      no extensions to core Prolog (like tabling, coroutines, constraints)
    \item
      slow, compared to good C implementations
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Mapping to low-level targets}
  \begin{itemize}
  \item
    Mechanically translate the interpreter to multiple
    lower-level targets
    \begin{itemize}
    \item C-like
    \item .NET
    \item Java
    \end{itemize}
  \pause
  \item
    Insert low-level aspects into the code as required by
    the target
    \begin{itemize}
    \item object layout
    \item memory management
    \end{itemize}
  \pause
  \item
    Optionally insert new pervasive features not expressed
    in the source
    \begin{itemize}
    \item continuations, ``micro-threads''
    \item dynamic compilation
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Translation Aspects (i)}
  Features not present in the source can be added during translation:
  \begin{itemize}
  \item
    \alert{memory management}: use different GC strategies (Boehm collector, custom mark-n-sweep)
  \item
    \alert{Stackless transformation}: allows program to control its stack (continuations, \dots)
  \end{itemize}
\end{frame}

@@

\section{The Prolog interpreter}
\subsection{Interpreter}
\begin{frame}
  \frametitle{Prolog Interpreter Implementation}
  \begin{itemize}
  \item
    naive, very simple interpreter
    \begin{itemize}
    \item uses "structure copying"
    \item interprets Prolog terms directly, no bytecode
    \end{itemize}
  \item
    uses continuation passing style inspired by BinProlog
  \item
    implements large parts of the ISO standard (some builtins missing)
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Interpreter Facts}
  \begin{itemize}
  \item
    2500 lines of Python code in total
  \item
    700 of those are for builtins
  \item
    after translation to C: 14000 lines of C code
  \item
    part of the PyPy distribution at:
    \texttt{http://codespeak.net/pypy}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
  \item
    Very simple Prolog interpreter in RPython can compete with interpreters on the JVM, CLR
  \item
    Interpreter implementation eased by use of a high-level language
  \item
    Low-level details abstracted away but re-inserted later
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Questions?}
  \begin{center}
  \includegraphics[scale=0.5]{image/py-web.png}

  \texttt{http://codespeak.net/pypy}
  \end{center}
\end{frame}
